`respR` - An R package for the analysis of respirometry data

### 2017-11-15 manuscript/LaTeX/manuscript-171115.Rmd
manuscript update

author:
- name       : Januar Harianto
  affiliation: University of Sydney, School of Medical Science, NSW Australia 2006
- name       : Nicholas Carey
  affiliation: Scottish Association for Marine Science, Oban, Argyll, UK, PA37 1QA
- name       : Maria Byrne
  affiliation: University of Sydney, School of Life & Environmental Sciences (SOLES), NSW Australia 2006

output:
  pdf_document:
    fig_caption    : yes
    keep_tex       : yes
    number_sections: yes
    latex_engine   : xelatex
    template       : template.tex
  html_document:
    fig_caption    : yes
    force_captions : yes
    highlight      : pygments
    number_sections: yes
    theme          : default

@@

keywords: "R, software, respirometry, metabolic rate"
thanks: "Replication files are available on the first author's Github account (http://github.com/januarharianto). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: nicholas.carey@sams.ac.uk."

@@

Data should be formatted correctly before use in `respR`. The function `inspect.data()` extracts a two-column data frame from a multi-column dataset and performs error checks on a data frame of any size, while plotting the data for quick data visualisation. All data should be numeric; respR at this time has limited support for date/time data (e.g. `POSIXct` and `POSIXt` classes). Parsing  different time formats is extremely complex and beyond the scope of this package, particularly because such packages already exist (e.g. `lubridate`). Time data, being a continuous variable, should be sequential, without duplicates. Ideally, sampling frequency will be monotonic, but in many cases this may not be the case, such as experiments with irregular sampling intervals, or because machine drift and/or data dropouts frequently occur during regular sampling. Therefore, `respR` uses time-based data extraction methods by default in most of its analytical functions (except `pcrit()`), and correctly handles irregular timeseries data.

Using `inspect.data()` is optional -- the main functions in our package will readily work with data frames that were not processed by the function. Running `inspect.data()` is an exploratory step that flags potential issues before analysis. Error check results are summarised in the R console and all tests are saved in the output object with the locations (i.e. row numbers) of the errors:

@@

```{r}
inspect.data(squid.rd, plot = F)
```

@@

Background oxygen consumption rate, $RO_{2(bg)}$, is often measured to account for the contribution of microbial respiration, and applied as a correction to experimental rates [@rogers_new_2016]. Background rates typically account for a small percentage of experimental rates, and multiple “blank” experiments are usually conducted and the rates averaged to obtain a more accurate estimate of the correction [e.g. @carey_size_2014; @daoud_temperature_2007]. The function `calc.rate.bg()` uses simple linear regression to simultaneously process multiple background rate measurements. Data segments can be truncated before analysis by time period or row numbers. The results are stored as an object for later use.

@@

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ [e.g. @gordon_energetics_1989; @chu_ecophysiological_2017]. The function `calc.rate()` can manually extract and process data subset from respiromety data. Data subsets can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accommodate most, if not all, manual data selection requirements and allow for consistent reporting of methods and results. Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results.

For more complex respirometry parameters, the function `auto.rate()` uses novel rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs regressions of sample size $n$ iteratively across the length of the data (i.e. rolling regressions), one sample at a time (**Fig. 2a**). Thus, a total of $(m - n) + 1$ number of overlapping regressions are fitted, which are then ranked to obtain maximum and minimum values. If an interval-based regression is required, we select the `"interval"` method when calling the function, and it automatically selects non-overlapping sections of the data for regressions (**Fig. 2b**).

```{r Figure2, echo = F, fig.cap="Illustration showing the sampling method used by `auto.rate()` during **(A)** interval regression analysis, where sample windows do not overlap and **(B)** rolling regression analysis, where sample windows overlap and move forward by one sample unit at a time."}

@@

The function `calc.rate.ft()` is used to specifically analyse flow-through respirometry data, and is similar to `calc.rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate.`calc.rate.ft()` can process data to determine the mean or modal $RO_2$ over a period.

```{r Figure3, echo = F, fig.cap="Typical graphical outputs of the function `auto.rate()`. **(A)** Depending on the `method` input, the function is able to show: (clockwise form top left) the maximum rate detected, the minimum rate detected, calculation of rate by fixed-width intervals, and the most linear region as determined by kernel density estimates. **(B)** In addition, the outputs will also print diagnostic plots of: (clockwise from top left) the rolling regression of rate as a function of time with dotted lines indicating the region used to estimate the parameter, a kernel density plot that shows only when the `linear` method is called, a residual plot of the regression, and a normal Q-Q plot of the residuals of the regression. Analysis is performed using the example dataset `sardine.rd`, which is included in the package. Code to reproduce the plots is available in the Supplementary Material."}

@@

As $P_{crit}$ is the break-point in the relationship between ambient dissolved oxygen ($DO_2$) and $RO_2$/$MO_2$, the user has two options to provide the required data. Loading an $O_2$ timeseries data into the function will prompt it to automatically perform a rolling mean and a rolling regression of $DO_2$ to extract those variables (**Fig. 4**). Alternatively, the user already has the required data, and the function may run the data directly using the argument `has.rate = TRUE`.

@@

For deliberate reasons, units of measurement are not used when determining $RO_2$ in `respR`. Instead, two functions, `convert.rate()` and `scale.rate()`, are used to transform $RO_2$ to specific units and perform volumetric or mass-specific conversions . This unique approach has the advantage of allowing "dimensionless" rate calculations for data that may not be fully supported by the entire respR workflow. For example, measurements of respiration can be scaled to surface area (e.g. diatoms, Iversen & Ploug, 2013, Naumann et al., 2014), volume (e.g. copepod eggs, Hammervold et al., 2015), or density (e.g. sea urchin larvae, Stumpp et al., 2011), all of which are currently not supported in `respR`. Users may also be interested in using respR's analytical functions to investigate other linear relationships or perform breakpoint analyses on non-respirometry-related data.

The function `convert.rate()` is used to convert units of oxygen concentration only (e.g. $mg/L$ or %) and has support for various unit combinations (Table 2). For ease of use, we have implemented a simple fuzzy string-matching algorithm to match relevant units even when they have been provided in different formats (e.g. "$mg/L$", "$mg/l$", "$mg L^{-1}$" and "$mg l^{-1}$" are identified as the same units). The function `scale.rate()` is conceptually similar to `convert.rate()`, but it focuses volumetric and mass-specific conversions (e.g. $mg h^{-1} kg^{-1}$).

@@

pander(conversions, caption = "Units available for conversions in `calc.rate()` and `scale.rate()`.")

>>>>

author:
- name       : Januar Harianto
  affiliation: University of Sydney, School of Medical Science, NSW Australia 2006
- name       : Nicholas Carey
  affiliation: Scottish Association for Marine Science, Oban, Argyll, UK, PA37 1QA
- name       : Maria Byrne
  affiliation: University of Sydney, School of Medical Science and School of Life & Environmental Sciences (SOLES), NSW Australia 2006

output:
  pdf_document:
    fig_caption    : yes
    keep_tex       : yes
    number_sections: yes
    latex_engine   : xelatex
    highlight      : tango
    template       : template.tex
  html_document:
    fig_caption    : yes
    force_captions : yes
    highlight      : seashell
    number_sections: yes
    theme          : default

@@

keywords: "R, software, respirometry, ecophysiology, metabolic rate"
thanks: "Replication files are available on the first author's Github account (http://github.com/januarharianto). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: nicholas.carey@sams.ac.uk."

@@

Data should be formatted correctly before use in `respR`. The function `inspect_data()` extracts a two-column data frame from a multi-column dataset and performs error checks on a data frame of any size, while plotting the data for quick data visualisation. All data should be numeric; respR at this time has limited support for date/time data (e.g. `POSIXct` and `POSIXt` classes). Parsing  different time formats is extremely complex and beyond the scope of this package, particularly because such packages already exist (e.g. `lubridate`). Time data, being a continuous variable, should be sequential, without duplicates. Ideally, sampling frequency will be monotonic, but in many cases this may not be the case, such as experiments with irregular sampling intervals, or because machine drift and/or data dropouts frequently occur during regular sampling. Therefore, `respR` uses time-based data extraction methods by default in most of its analytical functions (except `pcrit()`), and correctly handles irregular timeseries data.

Using `inspect_data()` is optional -- the main functions in our package will readily work with data frames that were not processed by the function. Running `inspect_data()` is an exploratory step that flags potential issues before analysis. Error check results are summarised in the R console and all tests are saved in the output object with the locations (i.e. row numbers) of the errors:

@@

```{r}
inspect_data(squid.rd, plot = F)
```

@@

Background oxygen consumption rate, $RO_{2(bg)}$, is often measured to account for the contribution of microbial respiration, and applied as a correction to experimental rates [@rogers_new_2016]. Background rates typically account for a small percentage of experimental rates, and multiple “blank” experiments are usually conducted and the rates averaged to obtain a more accurate estimate of the correction [e.g. @carey_size_2014; @daoud_temperature_2007]. The function `calc_rate_bg()` uses simple linear regression to simultaneously process multiple background rate measurements. Data segments can be truncated before analysis by time period or row numbers. The results are stored as an object for later use.

@@

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ [e.g. @gordon_energetics_1989; @chu_ecophysiological_2017]. The function `calc_rate()` can manually extract and process data subset from respiromety data. Data subsets can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accommodate most, if not all, manual data selection requirements and allow for consistent reporting of methods and results. Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results.

For more complex respirometry parameters, the function `auto_rate()` uses novel rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs regressions of sample size $n$ iteratively across the length of the data (i.e. rolling regressions), one sample at a time (**Fig. 2a**). Thus, a total of $(m - n) + 1$ number of overlapping regressions are fitted, which are then ranked to obtain maximum and minimum values. If an interval-based regression is required, we select the `"interval"` method when calling the function, and it automatically selects non-overlapping sections of the data for regressions (**Fig. 2b**).

```{r Figure2, echo = F, fig.cap="Illustration showing the sampling method used by `auto_rate()` during **(A)** interval regression analysis, where sample windows do not overlap and **(B)** rolling regression analysis, where sample windows overlap and move forward by one sample unit at a time."}

@@

The function `calc_rate_ft()` is used to specifically analyse flow-through respirometry data, and is similar to `calc_rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate.`calc_rate_ft()` can process data to determine the mean or modal $RO_2$ over a period.

```{r Figure3, echo = F, fig.cap="Typical graphical outputs of the function `auto_rate()`. **(A)** Depending on the `method` input, the function is able to show: (clockwise form top left) the maximum rate detected, the minimum rate detected, calculation of rate by fixed-width intervals, and the most linear region as determined by kernel density estimates. **(B)** In addition, the outputs will also print diagnostic plots of: (clockwise from top left) the rolling regression of rate as a function of time with dotted lines indicating the region used to estimate the parameter, a kernel density plot that shows only when the `linear` method is called, a residual plot of the regression, and a normal Q-Q plot of the residuals of the regression. Analysis is performed using the example dataset `sardine.rd`, which is included in the package. Code to reproduce the plots is available in the Supplementary Material."}

@@

As $P_{crit}$ is the break-point in the relationship between ambient dissolved oxygen ($DO_2$) and rate of O2 uptake (e.g. $RO_2$), the user has two options to input data. With an $O_2 \sim$ timeseries dataset the function will automatically perform a rolling mean $RO_2$ and a rolling regression of $DO_2$ (Fig. 4) in order to calculate $P_{crit}$ Alternatively, the user can input already calculated $RO_2/MO_2 \sim O_2$ data and the function will determine $P_{crit}$ using these data directly.

@@

For deliberate reasons, units of measurement are not used when determining $RO_2$ in `respR`. Instead, two functions, `convert_DO()` and `convert_rate()`, are used to transform $RO_2$ to specific units and perform volumetric or mass-specific conversions . This unique approach has the advantage of allowing "dimensionless" rate calculations for data that may not be fully supported by the entire respR workflow. For example, measurements of respiration can be scaled to surface area [e.g. diatoms & coral, @iversen_temperature_2013; @naumann_species-specific_2014], volume [e.g. copepod eggs, @hammervold_new_2015], or density [e.g. sea urchin larvae, @stumpp_co2_2011], all of which are currently not supported in `respR`. Users may also be interested in using respR's analytical functions to investigate other linear relationships or perform breakpoint analyses on non-respirometry-related data.

The function `convert_DO()` is used to convert units of oxygen concentration only (e.g. $mg/L$ or %) and has support for various unit combinations (Table 2). For ease of use, we have implemented a simple fuzzy string-matching algorithm to match relevant units even when they have been provided in different formats (e.g. "$mg/L$", "$mg/l$", "$mg L^{-1}$" and "$mg l^{-1}$" are identified as the same units). The function `convert_rate()` is conceptually similar to `convert_DO()`, but it focuses volumetric and mass-specific conversions (e.g. $mg h^{-1} kg^{-1}$).

@@

pander(conversions, caption = "Units available for conversions in `convert_DO()` and `convert_rate()`.")
